TacticToe: Learning to Prove with Tactics

### 2018-08-15 review_response.tex

We thank the reviewers for the valuable comments and questions. We have 
attempted to correct all the issues suggested in the reviews and we give 
further answers to the questions raised in this response letter.

\section*{Reviewer 1}

\begin{leftbar}
No data is provided to help other researchers reproduce this experiment.  I 
consider this is a flaw of the paper.  I think the paper should be improved in 
the direction of making the experiment reproducible by other researchers, as 
much as possible. The paper provides numerical evaluation of its results, 
mostly in the form of 
percentage of success under a given timeout. I would also be interesting to 
know how much it takes to re-run the whole process of learning and then using 
for the complete database.
\end{leftbar}

A paragraph was added in the methodology subsection where we talk about the 
reproducibility of our results. We mention that TacticToe is part of the HOL4 
(development) distribution and give the location of its directory.
The time it would take to re-do the final experiment independently is about 3 
days on a single CPU.

@@

\section*{Reviewer 2}

\begin{leftbar}
The only reservation I have with the paper is that it is not clear to me 
whether and how the proof tool is able to generate two commonly-used tactics. 
Firstly, generating witnesses to existential goals using \texttt{EXISTS\_TAC 
tm}. If these 
tactics are out of scope for the proof tool, then are all goals containing 
existential
quantifiers impossible for the tool to prove? Secondly, reducing a goal using 
an implication theorem using \texttt{MATCH\_MP\_TAC thm}. Often such 
implication theorems 
are
highly specific to a single theory or even theorem, so how can tactics like 
this be effectively learned? In my opinion the paper would be improved by a 
brief
discussion on how these tactics could be learned, either in the present or a 
future system.
\end{leftbar}

The is able to predict this tactics with arguments fixed. We also tested a way 
to predict the arguments independently taking account that our predictor is 
hardly accurate by being able to try multiple arguments.

### 2018-08-15 tactictoe_jv.tex
Other types of arguments: wip

\paragraph{Abstraction of Tactic Arguments}
The first step is to abstract arguments of tactics by replacing them by a
placeholder, creating a tactic with an unspecified argument.
Suppose we have a recorded tactic $t$. Since $t$ is a \sml code tree, we can
try to abstract any of the
\sml subterms. Let $u$ be a \sml subterm of $t$, and $h$ a variable (playing
the role of a placeholder), we
can replace $u$ by $h$ to create an abstracted tactic.
We denote this intermediate tactic $t[h/u]$. By repeating the
process, multiple
arguments may be abstracted in the same tactic. Ultimately, the more
abstractions are performed the more general a tactic is, as many tactics
become instances of the same abstracted tactic. As a consequence, it becomes
harder and harder to predict suitable arguments. In our experiments, we create
one abstraction $\hat{t}$ for each tactic $t$ by abstracting all subterms of
type theorem list in $t$.

\paragraph{Instantiation of an Abstracted Tactic}
An abstracted tactic is not immediately applicable to a goal, since it contains
unspecified arguments. To apply an abstracted tactic $\hat{t}$, we first need
to
instantiate the placeholders $h$ inside $\hat{t}$. Because it is difficult to
build a general predictor effective on each type of argument, we manually
design different argument predictors for each type. Those predictors are
given a goal $g$ as input and return the best predicted argument for this
goal.

Our default algorithm relies on argument predictions for theorems with a
radius of 16. It uses the produced list
$\mathit{Predictions}^{\mathit{tactic}}_{16}(g)$ to replace all the
placeholders
in $\hat{t}$.
The type of theorem lists is a very common type in \holfour proofs and theorems contain
enough information to be predicted accurately.

@@

\paragraph{Selection of Abstracted Tactics}
As abstracted tactics do not appear in human proofs, we need to find a way to
predict them so that they can contribute to the \tactictoe proof search.
A straightforward idea is to try $\hat{t}$ before $t$ during the proof search.
However, this risks doing unnecessary work as the two may perform similar steps.
Therefore, we would like to decide
beforehand if one is better than the other.
In fact, we can re-use the orthogonalization module to do this task for us.
We add $\hat{t}$ to the competition, initially giving it the coverage of $t$.
If $\hat{t}$
wins, it is associated with $g$ and is included in the database of tactic
features and thus can be predicted during proof search.
After many orthogonalization competitions, the coverage of $\hat{t}$ may exceed
the coverage of
$t$. At
this point, the coverage of $\hat{t}$ is estimated on its own and not inherited
from
$t$ anymore.

@@

\paragraph{Reproducibility}
The \tactictoe project is part of the \holfour distribution 
which can be downloaded at \url{https://github.com/HOL-Theorem-Prover/HOL}.
The \tactictoe modules can be found in the directory
\url{HOL/src/tactictoe}. The README explains how to launch the recording 
process (which includes learning techniques such as orthogonalization),
use \tactictoe interactively and run a re-proving experiments on a theory. The 
recording process takes less than eight hours. An upper bound for 
the time necessary for the final experiment can be given by the formula:
\[\mathit{number\ of\ theorems} \times \mathit{timeout} = 7164 \times 60\ 
\mathit{seconds} = 
4.975\ \mathit{days}\] which translate in practice
to about $2.5\ \mathit{days}$ because successful searches stop before reaching 
the timeout. And finally, by running this experiment on 20 cores, we cut down 
the actual total time requirement to a few hours.

>>>>

\subsection{Abstraction of Tactic Arguments}
The first step is to abstract arguments of tactics by replacing them by a
placeholder, creating a tactic with an unspecified argument.
Suppose we have a recorded tactic $t$. Since $t$ is a \sml code tree, we can
try to abstract any of the
\sml subterms. Let $u$ be a \sml subterm of $t$, and $h$ a variable (playing
the role of a placeholder), we
can replace $u$ by $h$ to create an abstracted tactic.
We denote this intermediate tactic $t[h/u]$. By repeating the
process, multiple
arguments may be abstracted in the same tactic. Ultimately, the more
abstractions are performed the more general a tactic is, as many tactics
become instances of the same abstracted tactic. As a consequence, it becomes
harder and harder to predict suitable arguments. In our experiments, we create
one abstraction $\hat{t}$ for each tactic $t$ by abstracting all subterms of
type theorem list in $t$.

\subsection{Instantiation of an Abstracted Tactic}
An abstracted tactic is not immediately applicable to a goal, since it contains
unspecified arguments. To apply an abstracted tactic $\hat{t}$, we first need
to
instantiate the placeholders $h$ inside $\hat{t}$. Because it is difficult to
build a general predictor effective on each type of argument, we manually
design different argument predictors for each type. Those predictors are
given a goal $g$ as input and return the best predicted argument for this
goal. We discuss here specific algorithm for instantiating three common types 
of arguments: theorem list, theorem, term.

\paragraph{Theorem List}
A typical example for

Our default algorithm relies on argument predictions for theorems with a
radius of 16. It uses the produced list
$\mathit{Predictions}^{\mathit{tactic}}_{16}(g)$ to replace all the
placeholders
in $\hat{t}$.
The type of theorem lists is a very common type in \holfour proofs and theorems 
contain enough information to be predicted accurately.
We are helped by the fact that the order of theorems rarely matters in this 
type of tactics and

\paragraph{Theorem}

\paragraph{Term}

We re-use the scheme describe previously for changing the prediction of one 
argument into the prediction of a list of arguments.
The only things that changes is that we have to design a special predictor for 
terms. In this context we think that the predictions should come from the list 
of subterms present in the goal as a first approximation. So all algorithm will 
try

Term synthesis and instantiation is vast problem for higher-order theorems 
provers such as Satallax and inductive programming Hipsec.

Side experiments with independent term argument predictions 
shows that trying to instantiate the tactic with subterms from the goal improve 
the success rate by another 4 percent. 
Nevertheless, in the evaluation presented in this paper, the 
abstraction/instantiation 
method is evaluated only for list of 
theorems, tactics with other type of arguments were predicted with their 
original arguments.

@@

\subsection{Selection of Abstracted Tactics}
As abstracted tactics do not appear in human proofs, we need to find a way to
predict them so that they can contribute to the \tactictoe proof search.
A straightforward idea is to try $\hat{t}$ before $t$ during the proof search.
However, this risks doing unnecessary work as the two may perform similar steps.
Therefore, we would like to decide
beforehand if one is better than the other.
In fact, we can re-use the orthogonalization module to do this task for us.
We add $\hat{t}$ to the competition, initially giving it the coverage of $t$.
If $\hat{t}$
wins, it is associated with $g$ and is included in the database of tactic
features and thus can be predicted during proof search.
After many orthogonalization competitions, the coverage of $\hat{t}$ may exceed
the coverage of
$t$. At
this point, the coverage of $\hat{t}$ is estimated on its own and not inherited
from
$t$ anymore.

@@

\paragraph{Reproducibility}
The \tactictoe project is part of the \holfour distribution 
which can be downloaded at \url{https://github.com/HOL-Theorem-Prover/HOL}.
\tactictoe modules can be found in the directory
\url{HOL/src/tactictoe}. The \url{README} file explains how to launch the 
recording 
process (which includes learning techniques such as orthogonalization),
use \tactictoe interactively and run a re-proving experiments on a theory. The 
recording process takes less than eight hours. An upper bound for 
the time necessary for the final experiment can be given by the formula:
\[\mathit{number\ of\ theorems} \times \mathit{timeout} = 7164 \times 60\ 
\mathit{seconds} = 
4.975\ \mathit{days}\] which translate in practice
to about $2.5\ \mathit{days}$ because successful searches stop before reaching 
the timeout. And finally, by running this experiment on 20 cores, we cut down 
the actual total time requirement to a few hours.
